SimTrans: A Self-supervised Method for Biomedical Image Segmentation

### 2023-06-19 cvpr_latex/PaperForReview.tex
add equation and modify formatting

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}

@@

To address the scarcity of biomedical image datasets, researchers aim to develop models that can learn features directly from the data without the need for labeled data. This approach, known as self-supervised learning, allows models to learn features from the data itself. One widely studied self-supervised learning framework in the field is SimCLR, proposed by Hinton et al.
SimCLR consists of four main components: data augmentation, a base encoder, a projection head, and a contrastive loss. The framework is designed to learn powerful representations from unlabeled data. In SimCLR, data augmentation is applied to generate multiple augmented versions of each unlabeled sample. These augmented samples are then passed through the base encoder, which extracts their features. The projection head further processes the encoded features to produce informative representations. Finally, the contrastive loss function encourages similar representations of augmented samples from the same sample while pushing representations from different samples apart.

\begin{figure*}
  \begin{center}
  % \fbox{\rule{0pt}{2in} \rule{.9\linewidth}{0pt}}
  \includegraphics{Images/SimCLR.png}
  \caption{Framwork of SimCLR.}
  \end{center}
  \label{fig:simclr}
\end{figure*}

@@

The data augmentation module in SimCLR performs random transformations on given data samples to generate two correlated views of the same sample, denoted as  and . These two views are treated as a positive sample pair. Data augmentation techniques used in SimCLR include random cropping, resizing, color distortion, and Gaussian blur. These operations generate pseudo-labels, which are then utilized to train the model.
The encoder in SimCLR extracts feature vectors, represented as  and , from the augmented data samples. The choice of encoder is not constrained and commonly used backbone frameworks like VGG and ResNet can serve as encoders in SimCLR.
The projection head maps the data representations to the space of contrastive loss. A common approach is to use a single-layer multilayer perceptron (MLP) for this mapping, as shown in Equation (1):

Here,  represents the mapping function,  is the ReLU activation function. According to the experimental analysis by the authors of SimCLR, defining the contrastive loss on the projected feature representation  tends to yield better results compared to using the feature representation  directly from the encoder.
The contrastive loss module defines the contrastive loss function for the contrastive prediction task. The key component of SimCLR is the normalized temperature-scaled cross-entropy loss (NT-Xent) used as the contrastive loss, given by Equation (2):
First, it calculates the similarity between a positive pair and all negative representations. The cosine similarity function is commonly used as a distance metric. Then, negative cross-entropy loss is employed to calculate the loss function for the positive sample pair .

@@

Swin Transformer is a hierarchical structure that builds feature maps by merging image patches at deeper layers, allowing flexible modeling at multiple scales. It achieves a linear computational complexity relative to the image size by computing self-attention within each local non-overlapping window. Additionally, Swin Transformer enables cross-window connections, as shown in Figure 2. The self-attention computation of a new window spans the boundaries of previously processed windows in layer l, providing cross-window connections between adjacent non-overlapping windows. This operation has been found effective in tasks such as image classification, object detection, and semantic segmentation. The ability to model visual entities of various scales and replace quadratic complexity with linear computational complexity contributes to Swin Transformer's versatility as a backbone for various visual tasks.

\begin{figure*}
  \begin{center}
  % \fbox{\rule{0pt}{2in} \rule{.9\linewidth}{0pt}}
  \includegraphics{Images/ST_1.png}
  \end{center}
     \caption{Self-attention computation for non-overlapping windows.}
  \label{fig:st_1}
\end{figure*}

Swin Transformer consists of a Multi-head Self-Attention module (MSA) based on movable windows, followed by a two-layer Multilayer Perceptron (MLP) with Gaussian Error Linear Units (GELU) as the non-linear activation function. GELU is a high-performance neural network activation function that weights the inputs rather than gating them with signs as in ReLU. Layer Normalization (LN) is applied before each MSA module and each MLP, and a residual connection is employed after each module, as shown in Figure 3.
The computational complexity of the standalone MSA module in Swin Transformer has a quadratic relationship with the image size, as shown in Equation (3):
However, in Swin Transformer, the self-attention computation is performed within non-overlapping local windows, resulting in a linear computational complexity relative to the image size, as shown in Equation (4):
Here, M is a constant.
Furthermore, Swin Transformer incorporates relative position information into the attention computation, as shown in Equation (5):
Where Q, K, and V represent the query matrix, key matrix, and value matrix, respectively. d is the dimension, and B represents the relative position bias. Compared to using absolute positions, incorporating relative positional biases significantly improves the model's performance.
Additionally, Swin Transformer introduces inductive biases such as locality, hierarchical feature representation, and translation equivariance, enabling it to serve as a universal backbone network for various image recognition tasks.

\begin{figure*}
  % \begin{center}
  % \fbox{\rule{0pt}{2in} \rule{.9\linewidth}{0pt}}
  \includegraphics[width=0.5\linewidth]{Images/ST_2.png}
  % \end{center}
  \caption{Block of Swin Transformer.}
  \label{fig:st_2}
\end{figure*}

@@

The Residual Channel Attention (RCA) module combines channel attention and residual connections. By using channel attention, it can capture features from different channels and enhance channel-wise information. The residual connections then connect the features, allowing for more refined feature representation. The structure of the RCA module is illustrated in Figure 5.

% \begin{figure*}
%   \begin{center}
%   % \fbox{\rule{0pt}{2in} \rule{.9\linewidth}{0pt}}  
%   \includegraphics{Images/RCBA.png}
%   \end{center}
%      \caption{Structure of RCBA.}
%   \label{fig:rcba}
% \end{figure*}

@@

Multi-scale dilated convolution is composed of convolutions with different dilation rates. Firstly, the feature map undergoes multiple convolution layers with different dilation rates. Then, the output results from each convolution layer are cascaded together to form the output of the entire module. This approach not only addresses the issue of information discontinuity caused by dilation but also helps the model capture multi-scale feature information. Figure 6 illustrates dilated convolutions, with the left side showing a dilated convolution with a dilation rate of 2, and the right side showing a dilated convolution with a dilation rate of 3.

% \begin{figure*}
%   \begin{center}
%   % \fbox{\rule{0pt}{2in} \rule{.9\linewidth}{0pt}}
%   \includegraphics{Images/MSD.png}
%   \end{center}
%      \caption{Structure of MSD.}
%   \label{fig:msd}
% \end{figure*}

@@

We utilize Swin Transformer as the encoder in the SimCLR framework and train it using unlabeled data, allowing the Swin Transformer to learn the distribution and features of the data with the help of pseudo-labels. Then, we apply the learned feature encoder to downstream segmentation tasks. Based on SimCLR and Swin Transformer, we design a deep learning network model called SimTrans, as shown in Figure 4. In the downstream tasks, to achieve an end-to-end output, a simple decoder is employed. It fuses the features obtained at different stages of the Swin Transformer and includes a decoder module that upsamples the fused features for comparison with the label map. Additionally, to better train the model, we propose a weighted structured loss function based on existing loss functions. It is a weighted sum of binary cross-entropy loss (BCE) and intersection over union loss (IOU):
Here, Y represents the label, W represents the edge weight, and represents the average pooling operation.

\begin{figure*}
  \begin{center}
  % \fbox{\rule{0pt}{2in} \rule{.9\linewidth}{0pt}}
  \includegraphics[width=0.8\linewidth]{Images/SimTrans.png}
  \end{center}
     \caption{Structure of SimTrans.}
  \label{fig:simtrans}
\end{figure*}

@@

This study conducted experimental validation of the proposed deep learning model on the ATLAS (Automatic Labeling of Brain Stroke Lesion) dataset, which is a publicly available dataset for brain stroke lesion segmentation. The dataset consists of a total of 229 T1-weighted brain magnetic resonance imaging (MRI) images provided by 11 different medical research centers. These images are expertly annotated by professional physicians. Each three-dimensional image has a voxel size of 0.9mm × 0.9mm × 0.3mm and contains 189 slices. Each slice has dimensions of 233 × 197. The entire ATLAS dataset consists of 43,281 two-dimensional slices. In this study, 10,203 labeled data were processed and used for validation in the downstream segmentation task, while the remaining 33,078 slices were used for self-supervised learning to learn the features of brain stroke data.
Three evaluation metrics, namely Dice coefficient, mean intersection over union (IoU), and precision, were selected to quantify the performance of the proposed model.
All the experiments were conducted in the PyTorch framework and trained using Nvidia-RTX 3090Ti GPUs. The deep learning models were trained for a total of 50 epochs.

@@

\begin{figure*}
  \begin{center}
  % \fbox{\rule{0pt}{2in} \rule{.9\linewidth}{0pt}}
  \includegraphics{Images/DR_SimCLR.png}
  \end{center}
     \caption{Different representations of the same image produced by SimCLR.}
  \label{fig:dr_simclr}
\end{figure*}

@@

\begin{figure*}
  \begin{center}
  % \fbox{\rule{0pt}{2in} \rule{.9\linewidth}{0pt}}
  \includegraphics{Images/Vision.png}
  \end{center}
     \caption{Feature visualization.}
  \label{fig:vision}
\end{figure*}

Figure 8 displays the following images from left to right: the original brain MRI image, feature maps from the first stage, feature maps from the second stage, feature maps from the third stage, feature maps from the fourth stage, and the ground truth label map. From the visualization, it can be observed that the SimTrans model progressively increases its attention to the lesion regions, demonstrating its effectiveness in segmenting the target regions.
The comparison results between SimTrans and two other networks, ResUNet and ViT, are presented in Table 1.

\begin{table}
  \centering
  \begin{tabular}{lccc}
    \toprule
    Model & Dice & MIoU & Precision \\
    \midrule
    ResUNet & 0.6810 & 0.5552 & 0.5583 \\
    ViT & 0.6783 & 0.5496 & 0.5529 \\
    SimTrans & 0.7585 & 0.6424 & 0.6453 \\
    \bottomrule
  \end{tabular}
  \caption{Results.   Ours is better.}
  \label{tab:result_1}
\end{table}

@@

\begin{figure*}
  \begin{center}
  % \fbox{\rule{0pt}{2in} \rule{.9\linewidth}{0pt}}
  \includegraphics{Images/Results.png}
  \end{center}
     \caption{Model segmentation visualization.}
  \label{fig:result}
\end{figure*}

@@

\begin{table}
  \centering
  \begin{tabular}{lccc}
    \toprule
    Metric & ResUNet & ViT & SimTrans \\
    \midrule
    Parameters/MB & 24.01 & 108.25 & 87.32 \\
    Complexity/GMac & 15.24 & 55.42 & 47.51 \\
    \bottomrule
  \end{tabular}
  \caption{Results.   Ours is better.}
  \label{tab:result_2}
\end{table}

@@

List and number all bibliographical references in 9-point Times, single-spaced, at the end of your paper.
When referenced in the text, enclose the citation number in square brackets, for
example~\cite{Authors14}.
Where appropriate, include page numbers and the name(s) of editors of referenced books.
When you cite multiple papers at once, please make sure that you cite them in numerical order like this \cite{Alpher02,Alpher03,Alpher05,Authors14b,Authors14}.
If you use the template as advised, this will be taken care of automatically.

>>>>

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{cite}

@@

To address the scarcity of biomedical image datasets, researchers aim to develop models that can learn features directly from the data without the need for labeled data. This approach, known as self-supervised learning, allows models to learn features from the data itself. One widely studied self-supervised learning framework in the field is SimCLR\cite{SimCLR}, proposed by Hinton et al.
SimCLR consists of four main components: data augmentation, a base encoder, a projection head, and a contrastive loss. The framework is designed to learn powerful representations from unlabeled data. In SimCLR, data augmentation is applied to generate multiple augmented versions of each unlabeled sample. These augmented samples are then passed through the base encoder, which extracts their features. The projection head further processes the encoded features to produce informative representations. Finally, the contrastive loss function encourages similar representations of augmented samples from the same sample while pushing representations from different samples apart.

\begin{figure}[htbp]   %注意，这里设置是关键
	\centering
	\includegraphics[width=\linewidth,scale=1.00]{Images/SimCLR.png}
	\caption{Framwork of SimCLR.}
	\label{fig:simclr}
\end{figure}

@@

The data augmentation module in SimCLR performs random transformations on given data samples to generate two correlated views of the same sample, denoted as $x_i$ and $x_j$. These two views are treated as a positive sample pair. Data augmentation techniques used in SimCLR include random cropping, resizing, color distortion, and Gaussian blur. These operations generate pseudo-labels, which are then utilized to train the model.
The encoder in SimCLR extracts feature vectors, represented as $h_i$ and $h_j$, from the augmented data samples. The choice of encoder is not constrained and commonly used backbone frameworks like VGG and ResNet can serve as encoders in SimCLR.
The projection head maps the data representations to the space of contrastive loss. A common approach is to use a single-layer multilayer perceptron (MLP) for this mapping, as shown in Equation (1):

\begin{equation}
z_i=g\left(h_i\right)=W^{(2)} \sigma\left(W^{(2)} h_i\right)
\end{equation}

Here, $g(\cdot)$ represents the mapping function,  $\sigma$ is the ReLU activation function. According to the experimental analysis by the authors of SimCLR, defining the contrastive loss on the projected feature representation $z_i$ tends to yield better results compared to using the feature representation $h_i$ directly from the encoder.
The contrastive loss module defines the contrastive loss function for the contrastive prediction task. The key component of SimCLR is the normalized temperature-scaled cross-entropy loss (NT-Xent) used as the contrastive loss, given by Equation (2):

\begin{equation}
  \ell_{i, j}=-\log \frac{\exp \left(\operatorname{sim}\left(z_i, z_j\right) / \tau\right)}{\sum_{k=1}^{2 N} \mathbb{1}_{[k \neq i]} \exp \left(\operatorname{sim}\left(z_i, z_k\right) / \tau\right)}
\end{equation}
First, it calculates the similarity between a positive pair and all negative representations. The cosine similarity function is commonly used as a distance metric. Then, negative cross-entropy loss is employed to calculate the loss function for the positive sample pair $(i, j)$.

@@

Swin Transformer\cite{ST} is a hierarchical structure that builds feature maps by merging image patches at deeper layers, allowing flexible modeling at multiple scales. It achieves a linear computational complexity relative to the image size by computing self-attention within each local non-overlapping window. Additionally, Swin Transformer enables cross-window connections, as shown in Figure 2. The self-attention computation of a new window spans the boundaries of previously processed windows in layer $l$, providing cross-window connections between adjacent non-overlapping windows. This operation has been found effective in tasks such as image classification, object detection, and semantic segmentation. The ability to model visual entities of various scales and replace quadratic complexity with linear computational complexity contributes to Swin Transformer's versatility as a backbone for various visual tasks.

\begin{figure}[htbp]   %注意，这里设置是关键
	\centering
	\includegraphics[width=\linewidth,scale=1.00]{Images/ST_1.png}
	\caption{Self-attention computation for non-overlapping windows.}
	\label{fig:st_1}
\end{figure}

Swin Transformer consists of a Multi-head Self-Attention module (MSA) based on movable windows, followed by a two-layer Multilayer Perceptron (MLP) with Gaussian Error Linear Units (GELU) as the non-linear activation function. GELU is a high-performance neural network activation function that weights the inputs rather than gating them with signs as in ReLU. Layer Normalization (LN) is applied before each MSA module and each MLP, and a residual connection is employed after each module, as shown in Figure 3.
The computational complexity of the standalone MSA module in Swin Transformer has a quadratic relationship with the image size, as shown in Equation (3):

\begin{equation}
  \Omega(M S A)=4 h w C^2+2(h w)^2 C
\end{equation}

However, in Swin Transformer, the self-attention computation is performed within non-overlapping local windows, resulting in a linear computational complexity relative to the image size, as shown in Equation (4):

\begin{equation}
  \Omega(W-M S A)=4 h w C^2+2 M^2 h w C
\end{equation}

Here, $M$ is a constant.
Furthermore, Swin Transformer incorporates relative position information into the attention computation, as shown in Equation (5):

\begin{equation}
  \operatorname{Attention}(Q, K, V)=\operatorname{SoftMax}\left(\frac{Q K^T}{\sqrt{d}}+B\right) V
\end{equation}

Where $Q$, $K$, and $V$ represent the query matrix, key matrix, and value matrix, respectively. $d$ is the dimension, and $B$ represents the relative position bias. Compared to using absolute positions, incorporating relative positional biases significantly improves the model's performance.
Additionally, Swin Transformer introduces inductive biases such as locality, hierarchical feature representation, and translation equivariance, enabling it to serve as a universal backbone network for various image recognition tasks.

\begin{figure}[htbp]   %注意，这里设置是关键
	\centering
	\includegraphics[width=0.8\linewidth,scale=1.00]{Images/ST_2.png}
	\caption{Block of Swin Transformer.}
	\label{fig:st_2}
\end{figure}

@@

The Residual Channel Attention (RCA)\cite{RCAB} module combines channel attention and residual connections. By using channel attention, it can capture features from different channels and enhance channel-wise information. The residual connections then connect the features, allowing for more refined feature representation. The structure of the RCA module is illustrated in Figure 5.

\begin{figure}[htbp]   %注意，这里设置是关键
	\centering
	\includegraphics[width=\linewidth,scale=1.00]{Images/RCBA.png}
	\caption{Structure of RCBA.}
	\label{fig:rcba}
\end{figure}

@@

Multi-scale dilated convolution\cite{MSD} is composed of convolutions with different dilation rates. Firstly, the feature map undergoes multiple convolution layers with different dilation rates. Then, the output results from each convolution layer are cascaded together to form the output of the entire module. This approach not only addresses the issue of information discontinuity caused by dilation but also helps the model capture multi-scale feature information. Figure 6 illustrates dilated convolutions, with the left side showing a dilated convolution with a dilation rate of 2, and the right side showing a dilated convolution with a dilation rate of 3.

\begin{figure}[htbp]   %注意，这里设置是关键
	\centering
	\includegraphics[width=\linewidth,scale=1.00]{Images/MSD.png}
	\caption{Structure of MSD.}
	\label{fig:msd}
\end{figure}

@@

We utilize Swin Transformer as the encoder in the SimCLR framework and train it using unlabeled data, allowing the Swin Transformer to learn the distribution and features of the data with the help of pseudo-labels. Then, we apply the learned feature encoder to downstream segmentation tasks. Based on SimCLR and Swin Transformer, we design a deep learning network model called SimTrans, as shown in Figure 4. In the downstream tasks, to achieve an end-to-end output, a simple decoder is employed. It fuses the features obtained at different stages of the Swin Transformer and includes a decoder module that upsamples the fused features for comparison with the label map. Additionally, to better train the model, we propose a weighted structured loss function based on existing loss functions. It is a weighted sum of binary cross-entropy loss (BCE) and intersection over union loss (IOU):

\begin{equation}
  L(s, y)=w * L_{b c e}(s, y)+L_{i o u}(s, y)
\end{equation}

Here, $Y$ represents the label, $W$ represents the edge weight $w=1+5 *|(a p(x)-y)|)$, $\quad a p(\cdot)$ represents the average pooling operation.

\begin{figure}[htbp]   %注意，这里设置是关键
	\centering
	\includegraphics[width=\linewidth,scale=1.00]{Images/SimTrans.png}
	\caption{Structure of SimTrans.}
	\label{fig:simtrans}
\end{figure}

@@

This study conducted experimental validation of the proposed deep learning model on the ATLAS (Automatic Labeling of Brain Stroke Lesion) dataset\cite{ATLAS}, which is a publicly available dataset for brain stroke lesion segmentation. The dataset consists of a total of 229 T1-weighted brain magnetic resonance imaging (MRI) images provided by 11 different medical research centers. These images are expertly annotated by professional physicians. Each three-dimensional image has a voxel size of 0.9mm $\times$ 0.9mm $\times$ 0.3mm and contains 189 slices. Each slice has dimensions of 233 $\times$ 197. The entire ATLAS dataset consists of 43,281 two-dimensional slices. In this study, 10,203 labeled data were processed and used for validation in the downstream segmentation task, while the remaining 33,078 slices were used for self-supervised learning to learn the features of brain stroke data.
Three evaluation metrics, namely Dice coefficient, mean intersection over union (IoU), and precision, were selected to quantify the performance of the proposed model.
All the experiments were conducted in the PyTorch framework and trained using Nvidia-RTX 3090Ti GPUs. The deep learning models were trained for a total of 50 epochs.

@@

\begin{figure}[htbp]   %注意，这里设置是关键
	\centering
	\includegraphics[width=\linewidth,scale=1.00]{Images/DR_SimCLR.png}
	\caption{Different representations of the same image produced by SimCLR.}
  \label{fig:dr_simclr}
\end{figure}

@@

\begin{figure}[htbp]   %注意，这里设置是关键
	\centering
	\includegraphics[width=\linewidth,scale=1.00]{Images/Vision.png}
	\caption{Feature visualization.}
  \label{fig:vision}
\end{figure}

Figure 8 displays the following images from left to right: the original brain MRI image, feature maps from the first stage, feature maps from the second stage, feature maps from the third stage, feature maps from the fourth stage, and the ground truth label map. From the visualization, it can be observed that the SimTrans model progressively increases its attention to the lesion regions, demonstrating its effectiveness in segmenting the target regions.
The comparison results between SimTrans and two other networks, ResUNet\cite(ResUNet) and ViT\cite{ViT}, are presented in Table 1.

\begin{table}
  \centering
  \begin{tabular}{lccc}
    \toprule
    Model & Dice & MIoU & Precision \\
    \midrule
    ResUNet & 0.6810 & 0.5552 & 0.5583 \\
    ViT & 0.6783 & 0.5496 & 0.5529 \\
    SimTrans & 0.7585 & 0.6424 & 0.6453 \\
    \bottomrule
  \end{tabular}
  \caption{Segmentation results comparison.}
  \label{tab:result_1}
\end{table}

@@

\begin{figure}[htbp]   %注意，这里设置是关键
	\centering
	\includegraphics[width=\linewidth,scale=1.00]{Images/Results.png}
	\caption{Model segmentation visualization.}
  \label{fig:result}
\end{figure}

@@

\begin{table}
  \centering
  \begin{tabular}{lccc}
    \toprule
    Metric & ResUNet & ViT & SimTrans \\
    \midrule
    Parameters/MB & 24.01 & 108.25 & 87.32 \\
    Complexity/GMac & 15.24 & 55.42 & 47.51 \\
    \bottomrule
  \end{tabular}
  \caption{Parameter and Computational Complexity comparison.}
  \label{tab:result_2}
\end{table}

@@

List and number all bibliographical references in 9-point Times, single-spaced, at the end of your paper.
When referenced in the text, enclose the citation number in square brackets, for
example~.
Where appropriate, include page numbers and the name(s) of editors of referenced books.
When you cite multiple papers at once, please make sure that you cite them in numerical order like this.
If you use the template as advised, this will be taken care of automatically.
